---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Spatiotemporal Subsampling {#subsampling}

Despite the strengths of eBird data, species observations collected through citizen science projects present a number of challenges that are not found in conventional scientific data. In this chapter, we'll discuss three of these challenges: **spatial bias**, **temporal bias**, and **class imbalance**. Spatial and temporal bias refers to the tendency of eBird checklists to be distributed non-randomly in space and time, while class imbalance refers to fact that there will be many more non-detections than detections for most species. All three can impact our ability to make reliable inferences from eBird data.

<div class="exercise">
  <h2>Exercise</h2>
  Think of some examples of birder behavior that can lead to spatial and temporal bias. 
  
  <button class="solution">Solution</button>
<div class="solution-content">

- **Spatial bias:**  most eBirders sample near their homes, in easily accessible areas such as roadsides, or in areas and habitats of known high biodiversity.
- **Temporal bias:** eBirders preferentially sample when they are available, such as weekends, and at times of year when they expect to observe more birds, such as spring migration in North America.

</div>
</div>

Fortunately, all three of these challenges can largely be addressed by spatiotemporal subsampling of the eBird data prior to modeling. In particular, this consists of dividing space and time up into a regular grid (e.g. 5 km x 5 km x 1 week), and only selecting a subset of checklists from each grid cell. To deal with class imbalance, we can subsample detections and non-detections separately to ensure we don't lose too many detections.

For the spatial part of the subsampling we'll use the package `dggridR` to generate a regular hexagonal grid and assign points to the cells of this grid. Hexagonal grids are preferable to square grids because they exhibit significantly less spatial distortion.

## A toy example {#subsampling-toy}

To illustrate how spatial sampling on a hexagonal grid works, let's start with a simpe toy example. We'll generate 500 randomly placed points, construct a hexagonal grid with 5 km spacing, assign each point to a grid cell, then select a single point within each cell.

```{r subsampling-toy}
library(auk)
library(sf)
library(dggridR)
library(lubridate)
library(tidyverse)

# bounding box to generate points from
bb <- st_bbox(c(xmin = -0.1, xmax = 0.1, ymin = -0.1, ymax = 0.1), 
              crs = 4326) %>% 
  st_as_sfc() %>% 
  st_sf()
# random points
pts <- st_sample(bb, 500) %>% 
  st_sf(as.data.frame(st_coordinates(.)), geometry = .) %>% 
  rename(lat = Y, lon = X)

# contruct a hexagonal grid with ~ 5 km between cells
dggs <- dgconstruct(spacing = 5)
# for each point, get the grid cell
pts$cell <- dgGEO_to_SEQNUM(dggs, pts$lon, pts$lat)$seqnum

# sample one checklist per grid cell
pts_ss <- pts %>% 
  group_by(cell) %>% 
  sample_n(size = 1) %>% 
  ungroup()

# generate polygons for the grid cells
hexagons <- dgcellstogrid(dggs, unique(pts$cell), frame = FALSE) %>% 
  st_as_sf()
ggplot() +
  geom_sf(data = hexagons) +
  geom_sf(data = pts, size = 0.5) +
  geom_sf(data = pts_ss, col = "red") +
  theme_bw()
```

In the above plot, black dots represent the original set of 500 randomly placed points, while red dots represent the subsampled data, one point per hexagonal cell.

## Subsampling eBird data {#subsampling-ebird}

Now let's apply this same approach to the zero-filled American Flamingo data we produced in the [previous lesson](#presabs); however, now we'll temporally sample as well, at a resolution of one week, and sample presences and absences separately. We start by reading in the eBird data and assigning each checklist to a hexagonal grid cell and a week.

```{r subsampling-ebird-read}
# generate hexagonal grid with ~ 5 km betweeen cells
dggs <- dgconstruct(spacing = 5)
# read in data
ebird <- read_csv("data/ebird_amefla_zf.csv") %>% 
  # get hexagonal cell id and week number for each checklist
  mutate(cell = dgGEO_to_SEQNUM(dggs, longitude, latitude)$seqnum,
         year = year(observation_date),
         week = week(observation_date))
```

Now we sample a single checklist from each grid cell for each week, using `group_by()` to sample presences and absences separetely.

```{r subsampling-ebird-sample}
ebird_ss <- ebird %>% 
  group_by(species_observed, year, week, cell) %>% 
  sample_n(size = 1) %>% 
  ungroup()
```

<div class="exercise">
  <h2>Exercise</h2>
  How did the spatiotemporal subsampling affect the overall sample size as well as the prevalence of detections?
  
  <button class="solution">Solution</button>
<div class="solution-content">

```{r subsampling-ebird-sol}
# original data
nrow(ebird)
count(ebird, species_observed) %>% 
  mutate(percent = n / sum(n))

# after sampling
nrow(ebird_ss)
count(ebird_ss, species_observed) %>% 
  mutate(percent = n / sum(n))
```

```{r encounter-sss-pcts, include = FALSE}
pct_before <- count(ebird, species_observed) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(species_observed) %>% 
  pull(proportion) %>% 
  round(3) %>%
  scales::percent()
pct_after <- count(ebird_ss, species_observed) %>% 
  mutate(proportion = n / sum(n)) %>% 
  filter(species_observed) %>% 
  pull(proportion) %>% 
  round(3) %>%
  scales::percent()
```

So, the subsampling *decreased* the overall number of checklists from `r nrow(ebird)` to `r nrow(ebird_ss)`, but *increased* the prevalence of detections from `r pct_before` to `r pct_after`.

</div>
</div>