---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Applications {#applications}

Now that we know how to extract and zero-fill data from eBird, let's do something with these data! We'll start by summarizing the data to show the trajectory of observation frequency over the course of the year. Next, we'll make a basic presence-absence map of eBird observations. More advanced topics are covered in [part II](#model-intro) of this workshop, which focuses on using eBird data to model species distributions. We'll start a new script for this lesson: `02_applications.R`.

## Frequency trajectories {#applications-trajectory}

The genus *Cardellina* contains five New World warbler species, including some of the most spectacular birds in North America: [Canada Warbler](https://ebird.org/species/canwar/), [Red Warbler](https://ebird.org/species/redwar1/), and [Pink-headed Warbler](https://ebird.org/species/pihwar1/). Let's extract and zero-fill data for three of the species in this genus within Guatemala.

```{r applications-trajectory-extract}
library(auk)
library(sf)
library(rnaturalearth)
library(lubridate)
library(tidyverse)

f_ebd <- "data/ebd_cardellina.txt"
f_sed <- "data/sed_cardellina.txt"

# extract
filters <- auk_ebd("ebd_2014-2015_yucatan.txt", 
                   file_sampling = "ebd_sampling_2014-2015_yucatan.txt") %>% 
  auk_species(c("Canada Warbler", "Wilson's Warbler", 
                "Pink-headed Warbler")) %>% 
  auk_country("Guatemala") %>% 
  auk_complete() %>% 
  auk_filter(f_ebd, f_sed)

# zero-fill
cardellina_zf <- auk_zerofill(f_ebd, f_sed, collapse = TRUE)
count(cardellina_zf, scientific_name, species_observed)
```

Next, let's summarize these data, calculating the frequency of observation on eBird checklists by month.

```{r applications-trajectory-summarize}
cardellina_freq <- cardellina_zf %>% 
  mutate(month = month(observation_date)) %>% 
  group_by(scientific_name, month) %>% 
  summarize(obs_freq = mean(species_observed)) %>% 
  ungroup()
```

In preparation for plotting, we can add the common names of the species by joining in the `ebird_taxonomy` data frame, which is included with `auk`. In addition, we'll convert the integer month numbers to dates–using the midpoint of each month–to aid axis labeling.

```{r  applications-trajectory-common}
cardellina_comm <- cardellina_freq %>% 
  inner_join(ebird_taxonomy, by = "scientific_name") %>% 
  select(common_name, month, obs_freq) %>% 
  mutate(month_midpoint = ymd(str_glue("2019-{month}-15")))
```

Finally, let's make a frequency trajectory for these four species.

```{r applications-trajectory-plot}
ggplot(cardellina_comm) +
  aes(x = month_midpoint, y = obs_freq, color = common_name) +
  geom_point() +
  geom_line() +
  scale_x_date(date_breaks = "2 months",
               date_labels = "%b") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Month", y = "Observation Frequency",
       color = NULL) +
  theme(legend.position = "bottom")
```

We have several different patterns going on here:

- Pink-headed Warblers is a resident species, present year-round at fairly low abundance
- Wilson's Warbler spends the winter in Guatemala and is common during that period
- Canada Warbler appears to only pass through Guatemala during migration

## Maps {#applications-maps}

Continuing with the same dataset, let's make a presence-absence map for Pink-headed Warbler. We'll start by filtering the zero-filled data to only Pink-headed Warbler observations and converting these points to a spatial format using the `sf` package.

```{r applications-maps-sf}
pihwar <- cardellina_zf %>% 
  filter(scientific_name == "Cardellina versicolor") %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

Next, we'll use `rnaturalearth` to get some contextual GIS data to use in our maps. This is an amazing source for free GIS data for making maps.

```{r applications-maps-rne}
ne_country <- ne_countries(continent = "North America", returnclass = "sf") %>% 
  st_geometry()
ne_gt <- ne_countries(country = "Guatemala", returnclass = "sf") %>% 
  st_geometry()
# restrict to points falling within Guatemala, removes those over water
pihwar <- pihwar[ne_gt, ]
```

Finally, we'll make a presence-absence map, building it up in layers. There are lots excellent tools for mapping in R; however, here we'll use the basic `plot()` function from the `sf` package. Other good options include `ggplot2` and `tmap`.

```{r applications-maps-map}
par(mar = c(0.25, 0.25, 0.25, 0.25))
# start by defining the bounds of the map with an empty plot
plot(ne_gt, col = NA, border = NA)
# borders
plot(ne_country, col = "grey80", border = "white", add = TRUE)
plot(ne_gt, col = "grey70", border = "white", add = TRUE)
# not observed
pihwar_abs <- filter(pihwar, !species_observed) %>% 
  st_geometry()
plot(pihwar_abs, col = alpha("grey20", 0.3), pch = 19, cex = 0.25, add = TRUE)
# present
pihwar_pres <- filter(pihwar, species_observed) %>% 
  st_geometry()
plot(pihwar_pres, col = alpha("orange", 1), pch = 19, cex = 0.5, add = TRUE)
title("Pink-headed Warbler eBird Observations", line = -1)
legend("bottomright", 
       col = c("grey20", "orange"),
       legend = c("Not reported", "Present"),
       pch = 19)
box()
```

Based on this map, we see that Pink-headed Warbler is restricted to the southwestern highlands of Guatemala.

<div class="exercise">
  <h2>Exercise</h2>
  Make a similar map of Wilson's Warbler (*Cardellina pusilla*) observations.
  
  <button class="solution">Solution</button>
<div class="solution-content">

```{r applications-maps-sol}
# prepare data
wilwar <- cardellina_zf %>% 
  filter(scientific_name == "Cardellina pusilla") %>% 
  mutate(pres_abs = if_else(species_observed, "Present", "Not detected")) %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
wilwar <- wilwar[ne_gt, ]

# make map
par(mar = c(0.25, 0.25, 0.25, 0.25))
plot(ne_gt, col = NA, border = NA)
plot(ne_country, col = "grey80", border = "white", add = TRUE)
plot(ne_gt, col = "grey70", border = "white", add = TRUE)
# not observed
wilwar_abs <- filter(wilwar, !species_observed) %>% 
  st_geometry()
plot(wilwar_abs, col = alpha("grey20", 0.3), pch = 19, cex = 0.25, add = TRUE)
# present
wilwar_pres <- filter(wilwar, species_observed) %>% 
  st_geometry()
plot(wilwar_pres, col = alpha("orange", 1), pch = 19, cex = 0.5, add = TRUE)
title("Wilson's Warbler eBird Observations", line = -1)
legend("bottomright", 
       col = c("grey20", "orange"),
       legend = c("Not reported", "Present"),
       pch = 19)
box()
```

</div>
</div>