---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Occupancy {#occupancy}

In this lesson, we'll use occupancy models to estimate the occupancy of occupancy of Australian Ibis on eBird checklists in September in the Temperate and Subtropical Forest CMZ, while explicitly accounting for imperfect detection. First, we'll give a short presentation introducing occupancy modeling. The presentation can be downloaded in [PowerPoint](https://github.com/mstrimas/ebp-workshop/raw/master/12_occupancy.pptx) or [PDF](https://github.com/mstrimas/ebp-workshop/raw/master/12_occupancy.pdf) format, or [viewed on SpeakerDeck](https://speakerdeck.com/mstrimas/ebird-best-practices-ii-occupancy).

<iframe src="https://speakerdeck.com/player/7375bae455c04703a770c0c6ab26dc6a" width="700px" height="400px" frameborder="0" marginwidth="0" marginheight="0" scrolling="no" style="border:none;" allowfullscreen webkitallowfullscreen mozallowfullscreen></iframe>

Let's start by loading the packages and data required for this lesson.

```{r occupancy-data}
library(auk)
library(lubridate)
library(sf)
library(dggridR)
library(unmarked)
library(raster)
library(ebirdst)
library(MuMIn)
library(AICcmodavg)
library(fields)
library(tidyverse)
# resolve namespace conflicts
select <- dplyr::select
projection <- raster::projection

set.seed(1)

# ebird data
ebird <- read_csv("data/ebd_zf_sep_tst.csv") %>% 
  filter(common_name == "Australian Ibis") %>% 
  mutate(year = year(observation_date),
         # occupancy modeling requires an integer response
         species_observed = as.integer(species_observed))

# modis land cover covariates
habitat <- read_csv("data/pland-elev_location-year.csv") %>% 
  mutate(year = as.integer(year))

# combine ebird and modis data
ebird_habitat <- inner_join(ebird, habitat, by = c("locality_id", "year"))

# prediction surface
pred_surface <- read_csv("data/pland-elev_prediction-surface.csv")
# latest year of landcover data
max_lc_year <- pred_surface$year[1]
r <- raster("data/prediction-surface.tif")

# load gis data for making maps
map_proj <- st_crs(3577)
ne_land <- read_sf("data/gis-data.gpkg", "ne_country") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
cmz <- read_sf("data/gis-data.gpkg", "cmz") %>% 
  filter(cmz_name == "Eastern Australia Temperate and Subtropical forests") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
ne_state_lines <- read_sf("data/gis-data.gpkg", "ne_state_lines") %>% 
  st_transform(crs = map_proj) %>% 
  st_geometry()
```

## Data preparation {#occupancy-prep}

Since we'll be fitting a single-season occupancy models, we'll need to start by focusing on observations from September of a single year, in this case the most recent year for which we have data. At this point, we also suggest subsetting the data to observations with 5 or fewer observers since there are few checklists with more than 5 observers.

```{r occupancy-prep-filter}
# filter to a single year of data
ebird_filtered <- filter(ebird_habitat, 
                         number_observers <= 5,
                         year == max(year))
```

As we learned in [Part I](#advanced-unmarked) of this workshop, we can use the `auk` function `filter_repeat_visits()` to extract just the eBird records that are suitable for occupancy modeling. Specifically, we want repeat visits to the same location by the same observer, so we'll use latitude, longitude, and observer ID to define 'sites'.

```{r occupancy-prep-repeats, class.source="livecode"}
# subset for occupancy modeling
occ <- filter_repeat_visits(ebird_filtered, 
                            min_obs = 2, max_obs = 10,
                            annual_closure = TRUE,
                            date_var = "observation_date",
                            site_vars = c("latitude", "longitude", 
                                          "observer_id"))
```

<div class="exercise">
  <h2>Exercise</h2>
  Suppose you define the temporal period of closure as week long blocks, rather than the whole month of September. Use `filter_repeat_visits()` to extract eBird data accordingly.
  
  <button class="solution">Solution</button>
<div class="solution-content">

```{r occupancy-prep-repeats-sol}
occ_days <- filter_repeat_visits(ebird_filtered, 
                                 min_obs = 2, max_obs = 10,
                                 n_days = 7,
                                 date_var = "observation_date",
                                 site_vars = c("latitude", "longitude", 
                                               "observer_id"))
```

</div>
</div>

<div class="exercise">
  <h2>Exercise</h2>
  Subsetting eBird data to just the observations suitable for occupancy modeling will inevitably reduce the amount of data. What proportion of observations remain after calling `filter_repeat_visits()`?  How many unique sites do we have?
  
  <button class="solution">Solution</button>
<div class="solution-content">

```{r occupancy-prep-loss-sol}
nrow(occ) / nrow(ebird_habitat)
n_distinct(occ$site)
```

</div>
</div>

Next, we'll use the `auk` function `format_unmarked_occu()` to put data in the specific format required by `unmarked`, as discussed in [Part I](#advanced-unmarked). Prior knowledge of Australian Ibis, as well as the predictor importance results from the [encounter rate](#encounter-habitat) lesson, inform what land cover variables we choose as occupancy covariates. The five effort variables should always be included as detection covariates, but we'll also examine whether different habitat types affect the detection probability.

```{r occupancy-prep-unmarked, class.source="livecode"}
# format for unmarked, select occupancy and detection covariates
occ_wide <- format_unmarked_occu(occ, 
                                 site_id = "site", 
                                 response = "species_observed",
                                 site_covs = c("n_observations", 
                                               "latitude", "longitude", 
                                               # % urban
                                               "pland_13", 
                                               # % savanna
                                               "pland_09",
                                               # % evergreen broadleaf forest
                                               "pland_02",
                                               # % woody savanna
                                               "pland_08"),
                                 obs_covs = c("time_observations_started", 
                                              "duration_minutes", 
                                              "effort_distance_km", 
                                              "number_observers", 
                                              "protocol_type",
                                              "pland_13", 
                                              "pland_09"))
```

As described in lesson \@ref(subsampling), we'll use spatial subsampling to reduce spatial bias. However, here we'll subsample at the level of 'sites' rather than observations.

```{r encounter-prep-sss, results="hide"}
# generate hexagonal grid with ~ 5 km betweeen cells
dggs <- dgconstruct(spacing = 5)
# get hexagonal cell id for each site
occ_wide_cell <- occ_wide %>% 
  mutate(cell = dgGEO_to_SEQNUM(dggs, longitude, latitude)$seqnum)
# sample one checklist per grid cell
occ_ss <- occ_wide_cell %>% 
  group_by(cell) %>% 
  sample_n(size = 1) %>% 
  ungroup() %>% 
  select(-cell)
```

Finally, we'll convert this data frame of observations into an `unmarked` object in order to fit occupancy models.

```{r encounter-data-unmarked, class.source="livecode"}
# creat unmarked object
occ_um <- formatWide(occ_ss, type = "unmarkedFrameOccu")
```

## Occupancy modeling {#occupancy-model}

Now that the data are prepared, we can fit a single-season occupancy model to using the `occu()` function, specifying the detection and occupancy covariates, respectively, via a double right-hand sided formula of the form `~ detection covariates ~ occupancy covariates`.

```{r occupancy-model-fit, class.source="livecode"}
# fit model
occ_model <- occu(~ time_observations_started + 
                    duration_minutes + 
                    effort_distance_km + 
                    number_observers + 
                    protocol_type +
                    pland_02 + pland_08
                  ~ pland_13 + pland_09 + pland_02 + pland_08, 
                  data = occ_um)
# look at the regression coefficients from the model
summary(occ_model)
```

### Assessment {#occupancy-model-assess}

The MacKenzie and Bailey [-@mackenzieAssessingFitSiteoccupancy2004] goodness-of-fit test can be used to assess the occupancy model fit. Note that to produce accurate results, this process requires simulating about 1,000 bootstrap samples, which can take a long time to run. For the sake of speed, if you want to run the below code, we suggest using `nsim = 5`.

```{r occupancy-model-assess, eval=FALSE, echo=1:2, class.source="dontrun"}
occ_gof <- mb.gof.test(occ_model, nsim = 1000, plot.hist = FALSE)
print(occ_gof)
saveRDS(occ_gof, "raw-data/ausibi_occupancy-model_gof.rds")
```

```{r occupancy-model-assess-actual, echo=FALSE}
# read in saved gof test results 
occ_gof <- readRDS("raw-data/ausibi_occupancy-model_gof.rds")
# print without chisq table
occ_gof$chisq.table <- NULL
print(occ_gof)
```

## Prediction {#occupancy-predict}

Now we can estimate the distribution of Australian Ibis in the CMZ and produce a map. Recall that when we [predicted encouter rate](#encounter-predict), we had to include effort variables in our prediction surface. We don't need to do that here because the estimated occupancy doesn't depend on the effort covariates, these only occur in the detection submodel. In addition, `predict()` can produce both predictions as well as estimates of the standard error.

```{r occupancy-predict-predict, eval=FALSE, echo=1:10, class.source="livecode"}
# make prediction
occ_pred <- predict(occ_model, 
                    newdata = as.data.frame(pred_surface), 
                    type = "state")

# add to prediction surface
pred_occ <- bind_cols(pred_surface, 
                      occ_prob = occ_pred$Predicted, 
                      occ_se = occ_pred$SE) %>% 
  select(latitude, longitude, occ_prob, occ_se)

saveRDS(pred_occ, "raw-data/ausibi_occupancy-model_predictions.rds")
```

```{r occupancy-predict-predict-load, echo=FALSE}
pred_occ <- readRDS("raw-data/ausibi_occupancy-model_predictions.rds")
```

<div class="checkpoint">
  <h2>Checkpoint</h2>
  Predicting on the full prediction surface will typically take several minutes. As the above code runs, let's take a short break.
</div>

Next, we want to plot these predictions. We'll convert this data frame to spatial features using `sf`, then rasterize the points using the prediction surface raster template.

```{r occupancy-predict-rasterize}
r_pred <- pred_occ %>% 
  # convert to spatial features
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = projection(r)) %>% 
  # rasterize
  rasterize(r)
r_pred <- r_pred[[c("occ_prob", "occ_se")]]
```

Finally, we can map these predictions!

```{r occupancy-predict-map, fig.asp=1.6}
# project predictions
r_pred_proj <- projectRaster(r_pred, crs = map_proj$proj4string, method = "ngb")

par(mfrow = c(2, 1))
for (nm in names(r_pred)) {
  r_plot <- r_pred_proj[[nm]]
  
  par(mar = c(0.25, 0.25, 0.25, 5))
  # set up plot area
  plot(cmz, col = NA, border = NA)
  plot(ne_land, col = "#dddddd", border = "#888888", lwd = 0.5, add = TRUE)

  # occupancy probability or standard error
  if (nm == "occ_prob") {
    title <- "Australian Ibis Occupancy Probability"
    brks <- seq(0, 1, length.out = 21)
    lbl_brks <- seq(0, 1, length.out = 11) %>% 
      round(2)
  } else {
    title <- "Australian Ibis Occupancy Uncertainty (SE)"
    mx <- ceiling(1000 * cellStats(r_plot, max)) / 1000
    brks <- seq(0, mx, length.out = 21)
    lbl_brks <- seq(0, mx, length.out = 11) %>% 
      round(3)
  }
  pal <- abundance_palette(length(brks) - 1)
  plot(r_plot, 
       col = pal, breaks = brks, 
       maxpixels = ncell(r_plot),
       legend = FALSE, add = TRUE)
  
  # borders
  plot(ne_state_lines, col = "#ffffff", lwd = 0.75, add = TRUE)
  box()
  
  # legend
  par(new = TRUE, mar = c(0, 0, 0, 0))
  image.plot(zlim = range(brks), legend.only = TRUE, 
             breaks = brks, col = pal,
             smallplot = c(0.89, 0.92, 0.25, 0.75),
             horizontal = FALSE,
             axis.args = list(at = lbl_brks, labels = lbl_brks,
                              fg = "black", col.axis = "black",
                              cex.axis = 0.75, lwd.ticks = 0.5,
                              padj = 0),
             legend.args = list(text = title,
                                side = 2, col = "black",
                                cex = 1, line = 0))
}
```

## Model selection {#occupancy-select}

So far, we've been using a global model that includes all of the covariates we believe will influence the occupancy and detection probabilities; however, a more thorough approach is to use model selection to compare and rank a set of candidate models, each containing different subsets of the covariates. We'll use the `dredge()` function, which evaluates a set of candidate models generated by using different combinations of the terms in the global model. Since we know from prior experience that the effort covariates are almost always important, we'll lock these variables in, and consider a candidate set consisting of all possible combinations of the ecological covariates in the occupancy submodel.

```{r occupancy-model-select-dredge, class.source="livecode"}
# get list of all possible terms, then subset to those we want to keep
det_terms <- getAllTerms(occ_model) %>% 
  # retain the detection submodel covariates
  discard(str_detect, pattern = "psi")

# fit all possible combinations of the occupancy covariates
occ_dredge <- dredge(occ_model, fixed = det_terms)

# model comparison
select(occ_dredge, starts_with("psi(p"), df, AICc, delta, weight) %>% 
  mutate_all(~ round(., 3)) %>% 
  knitr::kable()
```

<div class="tip">
  <h2>Tip</h2>
  To determine which candidate model each row corresponds to, note that he columns beginning with `psi(` give the model coefficients for each of the habitat covariates in the occupancy submodel; missing values indicate that that covariate was not included in the given model.
</div>

The corrected [Akaike Information Criterion (AICc)](https://en.wikipedia.org/wiki/Akaike_information_criterion#AICc) measures the performance of each model, relative to the other models in the candidate set, adjusting for the number of parameters. Lower values indicate models with a better fit to the data, penalizing for the number of parameters. Delta is the difference between the AICc values for the given model and that for the top model (i.e. the one with the lowest AICc). Finally, the AIC weight is a transformation of delta that can be interpreted as the likelihood that the given model is the most likely one of the candidate models to have generated the data.

There doesn't appear to be a clear single model, or even a small set of models, that are most likely to have generated our data. Given this, we'll average across the top models (those comprising 95% of the weights), weighted by AICc, to produce a model-averaged prediction.

```{r occupancy-model-select-average, class.source="livecode"}
# select models with the most suport for model averaging
occ_dredge_95 <- get.models(occ_dredge, subset = cumsum(weight) <= 0.95)

# average models based on model weights 
occ_avg <- model.avg(occ_dredge_95, fit = TRUE)
```

Making model-averaged predictions works in the same way as for making [predictions from a single global model](#occupancy-predict). However, it takes significantly more time to run, so we won't run the below code or produce maps based on model-averaged predictions during the workshop, instead leaving it as an exercise.

```{r occupancy-model-select-predict, eval=FALSE, class.source="dontrun"}
# 
occ_pred_avg <- predict(occ_avg, 
                        newdata = as.data.frame(pred_surface), 
                        type = "state")

# add to prediction surface
pred_occ_avg <- bind_cols(pred_surface, 
                          occ_prob = occ_pred_avg$fit, 
                          occ_se = occ_pred_avg$se.fit) %>% 
  select(latitude, longitude, occ_prob, occ_se)
```

### Detection model {#occupancy-select-detection}

A unique feature of occupancy models, is that we can investigate whether certain covariates influence detection, which wasn't possible using the machine learning approach in lesson \@ref(encounter). We included evergreen broadleaf forest (`pland_02`) and woody savanna (`pland_08`) as detection covariates in the global model, and we can compare a set of models with and without these covariates to assess their importance.

```{r occupancy-select-detection-define}
# define occupancy model formula with only effort in detection submodel
det_mod <- ~ time_observations_started + 
  duration_minutes + 
  effort_distance_km + 
  number_observers + 
  protocol_type ~ 
  pland_13 + pland_09 + pland_02 + pland_08

# create new formulae with landcover covariates added to the detection submodel
mods <- list(det_mod_null = det_mod, 
             det_mod_for = update.formula(det_mod, ~ . + pland_02 ~ .),
             det_mod_ws = update.formula(det_mod, ~ . + pland_08 ~ .),
             global = update.formula(det_mod, 
                                     ~ . + pland_02 + pland_08 ~ .)) %>% 
  # fit candidate models
  map(occu, data = occ_um)
```

Now we can perform model selection on this set to compare the different candidate models.

```{r occupancy-select-detection-compare}
mod_sel <- fitList(fits = mods) %>% 
  modSel()
mod_sel
```

From these results, it's clear that including these habitat covariates (especially woody savanna) leads to an improvement in model performance, as shown by the AIC and AIC weights. Let's look at the coefficients from the global model for these covariates to see how they're impacting detection and occupancy. 

```{r occupancy-select-detection-coef}
coef(occ_model) %>% 
  enframe() %>% 
  filter(str_detect(name, "pland_"))
```

The `psi()` coefficients are from the occupancy submodel and the `p()` coefficients are from the detection submodel.

## Exercises {#occupancy-exercises}

Now that you've completed this lesson, try modifying your script to complete at least one of the following exercises:

1. Try sampling more than a single checklist per grid cell in the spatiotemporal sampling. How does that affect model fit and predictions?

2. What happens to the size of dataset if you only use stationary counts, or reduce the distance traveled to 1 km? How does it impact the results? How does the different input data affect your interpretation of the results? 

3. What happens to the size of the dataset if you allow repeat visits to be by multiple observers? How does this impact the results. 

4. Produce a map based on model averaged predictions. Note that making these predictions may take up to an hour.

5. Try estimating occupancy for one of the other species in the example data.

```{r occupancy-purl, eval=FALSE, echo=FALSE}
unlink("12_occupancy.R")
knitr::purl("12_occupancy.Rmd")
```
